Best practice examples for primary data publication

### 2019-03-27 manuscript/appendix_B.Rmd
update Appendix B, update affiliations, fix Example 2

---
title: "Best practice examples for primary data publication"
subtitle: "Towards an Ecological Trait-data Standard - Supplementary Material B"
author:
- name: Florian D. Schneider \footnotemark[1]  \footnotemark[2]
- name: David Fichtmüller \footnotemark[5]
- name: Martin M. Gossner \footnotemark[8]
- name: Anton Güntsch \footnotemark[5]
- name: Malte Jochum \footnotemark[4]
- name: Birgitta König-Ries\footnotemark[6]
- name: Gaëtane Le Provost \footnotemark[2]
- name: Pete Manning \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[6]
- name: Caterina Penone \footnotemark[4]
- name: Nadja K. Simons \footnotemark[7]
fontsize: 11pt
geometry: margin=1.2in
documentclass: scrartcl
csl: amnat.csl
bibliography: div05.bib
nocite: | 
  @gossner15a, @gossner15
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
---

@@

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}

@@

- **Use independent file formats!** For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma- or tab-separated text files  ('.csv' or '.tsv') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g. UTF-8 or ASCII). For internationalisation, use points as decimal delimiters and set date-formats to ISO 8601, i.e. "YYYY-MM-DD".

@@

- **Publish raw data!** When publishing trait-data, priority is on containing measurement values while minimizing the risk of purporting errors through data conversion or aggregation. Thus, data should be published as they have been recorded, i.e. in original units and raw values, along with standardised values that match a unified and compatible scheme of a data standard. This is true for the core terms, i.e. the trait value (the quality) and observation entity, but also for any co-variates reported along with it.

@@

A more effective method that allows for high flexibility and easier data merging is the storage of traitdata in long-table formats [@wickham14; @parr16]. 
This data format corresponds to the Entity-Quality Model: each row contains a single measurement or fact of a specific trait, referenced to a single specimen  (i.e., one occurrence)  which is assigned to a specific taxon  (see Table B2). 
Identifiers can provide structure and link measurements that are correlated by assigning a unique ID for each single individual. 
Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier. 
Long-table datasets purport multiple advantages for data handling (e.g. filtering, sub-setting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modelling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14]. 
As long-table data share a larger set of columns, merging of datasets is in most cases easier for the end-user.

@@

# 21228 life history traits of moths

@@

attr(moths, 'taxa') <- "species"
attr(moths, 'keep') <-  c(locationID = "plot")
attr(moths, 'individuals') <- "label"
attr(moths, 'units') <- c(weight = "mg", wing_length = "mm", wing_width = "mm", wing_area = "mm2", wing_loading = "mg mm-2")

@@

kable(moths[1:45,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
      booktabs = T, caption = caption) %>%
kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)

@@

mothsStd <- standardize(moths)
mothsStd <- mothsStd[order(as.numeric(substr(mothsStd$occurrenceID,2,10))),]

caption <- "Example 2 after standardisation into measurement long-table format. Each row contains a single measurement linked to an observation of an individual specimen, i.e. an occurrence, via identifiers."
select_cols <- c("scientificName","traitName","traitValue","traitUnit","traitID","scientificNameStd","taxonID","occurrenceID", "measurementID", "locationID")

kable(mothsStd[1:28,select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
      booktabs = T, caption = caption) %>%
kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)

>>>>

---
title: "Best practice examples for primary data publication"
subtitle: "Towards an Ecological Trait-data Standard - Supplementary Material B"
author:
- name: Florian D. Schneider \footnotemark[1]  \footnotemark[2]
- name: David Fichtmüller \footnotemark[3]
- name: Martin M. Gossner \footnotemark[4]
- name: Anton Güntsch \footnotemark[3]
- name: Malte Jochum \footnotemark[5] \hspace{1mm} \footnotemark[8]
- name: Birgitta König-Ries\footnotemark[6]
- name: Gaëtane Le Provost \footnotemark[2]
- name: Pete Manning \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[6]
- name: Caterina Penone \footnotemark[5]
- name: Nadja K. Simons \footnotemark[7] \hspace{1mm} \footnotemark[9]
fontsize: 11pt
geometry: margin=1.2in
documentclass: scrartcl
csl: amnat.csl
bibliography: div05.bib
nocite: | 
  @gossner15a, @gossner15
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
---

@@

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com,  \newline
c/o Birgitta König-Ries, Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Ernst-Abbe-Platz 2, 07743 Jena, Germany
}
\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
\footnotetext[3]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
\footnotetext[4]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
\footnotetext[5]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[8]{German Centre for Integrative Biodiversity Research (iDiv) Halle‐Jena‐Leipzig, Leipzig, Germany \& Leipzig University, Institute of Biology, Leipzig, Germany}
\footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}
\footnotetext[9]{Ecological Networks, Department of Biology, Technische Universität Darmstadt, Darmstadt, Germany}

@@

- **Use independent file formats!** For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats but rather in software-independent file formats (like .xlsx or .odt), or comma- or tab-separated text files ('.csv' or '.tsv') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g. UTF-8 or ASCII). For internationalisation, use points as decimal delimiters and set date-formats to ISO 8601, i.e. "YYYY-MM-DD".

@@

- **Publish raw data!** When publishing trait-data, priority is on containing measurement values while minimizing the risk of purporting errors through data conversion or aggregation. Thus, data should be published as they have been recorded, i.e. in original units and raw values, along with standardised values that match a unified and compatible scheme of a data standard. This is true for the core terms, i.e. the trait value and the item of observation and its taxon assignment, but also for any co-variates reported along with it.

@@

A more effective method that allows for high flexibility and easier data merging is the storage of traitdata in long-table formats [@wickham14; @parr16]. 
This data format corresponds has a row-based structure: each row contains a single measurement or fact of a specific trait, referenced to a single specimen  (or one occurrence of an individual organism at a particular space and time) which is assigned to a specific taxon  (see Table B2). 
Identifiers can provide structure and link measurements that are correlated by assigning a unique ID for each single individual. 
Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier. 
Long-table datasets purport multiple advantages for data handling (e.g. filtering, sub-setting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modelling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14]. 
As long-table data share a larger set of columns, merging of datasets is in most cases easier for the end-user.

@@

```{r,  echo = FALSE, warning = FALSE, message=FALSE}
library(XML)
library(RCurl)
library(getPass)

read.service <- function(datasetid, 
                         user = NULL, pswd = NULL, 
                         dec=".", 
                         na.strings="NA", 
                         fill=FALSE, sep="\t", 
                         quote=if(identical(sep, "\n")) "" else "'\"",
                         fileEncoding = "UTF-8")
{

  if(is.null(user)) user <- readline("user name: ") 
  if(is.null(pswd)) pswd <- getPass::getPass("password: ", noblank = FALSE)

  opts = curlOptions(encoding="CE_UTF8" ,ssl.verifypeer = FALSE, httpheader=c(Accept = "text/plain"))
  params = c("datasetId"=datasetid, "username"=user, "password"=pswd)
  data <- postForm(uri="https://www.bexis.uni-jena.de/WebServices/DataService.asmx/DownloadData", .params=params, .opts=opts, style="post")  
  x <- xmlTreeParse(data, asText = T, trim=FALSE)
  txt <- xmlValue(xmlRoot(x)[[1]])
  txt <-gsub("#","?", txt)
  f <- textConnection(toString(txt))   
  d <- utils::read.table(file=f, sep=sep, header=T, dec=dec, na.strings=na.strings, fill=fill, quote=quote, fileEncoding = fileEncoding)
  close(f)
  d
}
# 21228 life history traits of moths

@@

attr(moths, 'taxa') <- "species"
attr(moths, 'keep') <-  c(locationID = "plot")
attr(moths, 'occurrences') <- "label"
attr(moths, 'units') <- c(weight = "mg", wing_length = "mm", wing_width = "mm", wing_area = "mm2", wing_loading = "mg mm-2")

select_rows <- 1:12

@@

kable(rbind(moths[select_rows, select_cols]), longtable=TRUE, escape = TRUE, row.names = FALSE,
      booktabs = T, caption = caption, digits = c(NA,NA,NA,3, 2, 2,4,4)) %>%
kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)

@@

mothsStd <- standardize(moths)
mothsStd <- mothsStd[order(as.numeric(substr(mothsStd$occurrenceID,2,10))),]
names(mothsStd)[c(1,2,6,7)] <- c("verbatimScientificName", "verbatimTraitName", "scientificName", "traitName")

caption <- "Example 2 after standardisation into measurement long-table format. Each row contains a single measurement linked to an observation of an individual specimen, i.e. an occurrence, via identifiers."
select_cols <- c("scientificName","traitName","traitValue","traitUnit","traitID","verbatimTraitName","taxonID","occurrenceID", "measurementID", "locationID")
select_rows <- which(mothsStd$traitName %in% c("body_weight", "wing_area", "wing_length") )

kable(mothsStd[select_rows[1:28],select_cols], longtable=TRUE, escape = TRUE, row.names = FALSE,
      booktabs = T, caption = caption, digits = c(4)) %>%
kable_styling(latex_options = c("scale_down", "repeat_header", "striped"), font_size = 5)
